\normalfont\spacedallcaps{The economic consequences of different attitudes of a policy maker: a combined epidemiological-econometric study}

### 2021-03-24 Manuscript/manuscript.tex
PUB

Within a standard compartmental model for describing the dynamics of epidemic (Susceptible-Infectious-Recovered-Dead),  we considered a policy-maker (PM) that imposes stochastically different types of lock-downs. The probability that tunes this stochastic process reflects his/her different attitude to face an epidemics (e.g., \textit{laissez-faire} vs. very strict).  In order to simulate not only an epidemiological scenario but also an economic one, we estimated, via a Difference-in-Difference regression,  the impact of national lock-downs applied during the first wave of COVID19 in Italy from March 2020 to June 2020,  on two microeconomic sectors: sales value and redundancy funds (\textit{Cassa Integrazione}).  We found that by modifying with continuity the PM attitude to impose the lock-down, a phase transition (as defined for a physical system) is obtained.  The comparison of these two scenarios and their impact provides a bird's-eye view of the socio-economic consequences of the PM attitude

@@

The recent pandemic due to the spread of the SARS-CoV-2 virus opened a highly debated issue about the best approach for the policy-maker to face the epidemic.  Unlike the past pandemics (e.g., Spanish Flu,  Asiatic Flu, and Hong-Kong Flu), a massive amount of data are easily accessible for this pandemic.  Consequently, the modeling of the virus diffusion and the effect on socio-economic texture for different countries can be investigated with a more satisfactory resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally, the reproduction number, as described in Supporting Information).  Indeed the policy-maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme,  the spread of the virus is,  of course, stopped but,  on the other side,  the toll for such approach is that not only the economic activity (and so the income of people/firms) but also that the furniture of many primary goods is stopped.  On the other side,  if the PM takes no lock-down measures, the toll to be paid will be not only the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice, the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion to others (for this purpose, an excellent analysis was provided by Li et al. in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).
As a consequence, the lock-down efficacy,  within certain limits,  can be tuned.  In the literature, different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular, for the model in Ref. \cite{kabir2020evolutionary}, it is assumed that the PM can impose a continuous factor for the lock-down (e.g., he/she can reduce the daily number of contacts between people by choosing each value between a definite range). In contrast, for different countries, such a factor seems to be much more discrete (e.g., the PM can reduce the daily contacts only by choosing definite values). Moreover it is assumed that the policy-maker perfectly knows the consequences of his/her choices and that he/she can act without delay to impose the optimal choice: this feature is entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy-maker.  Here we would propose here an alternative way to model the decision of the policy-maker that is based on a stochastic model instead of on a deterministic one. Moreover we will consider a discrete intensity for the lock-down. Finally,  differently to the previous researches that focused basically on macroeconomic impact,  here we estimated and put in the model the impact of the different lock-downs at the microeconomic level: in particular, through the difference in difference,  we evaluated the effect of the different levels of lock-down on different sales sector as well as on redundancy funds (\textit{Cassa Integrazione Ordinaria}) in Italy.  Thus, the final output of the model will be the cumulative deaths, the economic damage for each sales sector and the the increase of redundancy funds paid.
Here we will also consider that there is not only an economic cost for each death as done by \cite{kabir2020evolutionary}, but there is also average cost for each infected person (referring to Italian data) because a consistent part of them should assisted by the national health system.  As we will show by varying with continuity the probabilistic parameter by which the PM imposes the lock-down,  a discontinuity in the epidemiological model was obtained.  Such behaviour belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase, we will discuss the result of the simulation, and then we will compare them to get a general insight.

\section{Model and methods}

The model of the present study is composed of an epidemiological part that shapes the diffusion of the virus. Then its output is used by the economic model to quantify the damage. Thus we will discuss the epidemiological part and then the economic one

@@

where S is the number of people that are still susceptible, I the number of people that are infected, R people that are recovered, and D are people that are dead.  N denotes the total population that for the timing of this paper it will be considered fixed\footnote{Otherwise if longer horizontal timing is considered, it is necessary to consider a source term for the births and a well term for the natural deaths. For further details see \cite{vynnycky2010introduction}}. On the other side, $\beta$,$\gamma$, and $\mu$ are the parameters that shape the probability by which one individual in the model moves from a compartment to another: in particular, $\beta$ is the probability to be infected, $\gamma $ the probability to recover and $\mu$ the probability to die (basically the lethality that is defined as the probability to die given to be ill).  Usually, epidemiologist is interested in the ratio:  

@@

known as the basic reproduction factor. This number is the average number of people that a single individual infects and describes if the epidemic is in negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in positive feedback (R$_{0}$>1).  As a consequence, if the epidemic is within a negative feedback will be dissipated, while if it is in positive feedback will grow.  Note that in this simple model, since the parameters are not time-dependent, this factor is constant.  As performed by Ferrari et al. \cite{ferrari2020modelling},  when time-dependent parameters $\beta$ and $\gamma$ are taken in to account the reproduction factor R$_{0}$ becomes time depending: thus scholar rename it as $R_{t}$.  For the present work,  we limited to constant parameters. In particular, we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$.  The $\mu$ was set in order to keep in to account the calculated lethality for Italy: 1 \% \cite{fernandez2020estimating} (see also the excellent analysis made by the Institute for International Political Studies (ISPI) \cite{ISPI})\footnote{The authors are aware that this ratio is far from being homogeneous for the different ages of the population: however if this factor is taken into account,  it requires the solution of a system of partial derivative equations. In this case, the numerical calculations become much more complicated}.  The overall population $N$ was set to 60M to simulate the Italian population.  Within the daily temporal evolution of this model, which was obtained by numerically solving the differential equation above via the \textit{DeSolve} package, we considered a trigger activated by the PM every seven days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM imposes laws that reduces the $\beta$ factor by a multiplicative factor equal to $0.7$ (and thus the reproduction factor $R_{0}$) if the normalized infected people are more than ten times the previous threshold he/she will impose with certain probability restrictions that reduce the $\beta$ factor by a multiplicative factor of $0.25$; finally if the threshold is exceeded more than 50 times, the PM will impose with a certain probability restrictive measures that reduce the $\beta$ a multiplicative factor equal to $0.025$. These attenuation parameters were adjusted, considering the results of Marziano et al. in Ref. \cite{marziano2021retrospective}.  Therefore such a trigger makes the  R$_{0}$ parameter time-dependent,  although in a discrete way.  As we said, the PM acts with a certain probability,  more formally stochastically; each week, a random number (from zero to one) is extracted: if this is higher than a certain threshold,  the relative restrictive decision is taken, otherwise not. The threshold value captures the PM attitude to impose the lock-down: lower values model a careful PM, high value a lazy one. In this way, the model can simulate different scenarios for the different PM attitudes: as we will see, this can produce two very different results.  At the end of the simulation, besides the values given by the standard SIRD model (recovered and deaths),  the algorithm also provides the number of weeks in which each restriction was active: these values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy. It is worth noting that here as lock-down, we considered only the national one applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modeling a unique system is easier concerning modeling an ensemble of communicating clusters that represent regions or provinces: therefore, if one is interested in modeling the second wave of pandemics, such an approach should be undertaken. Moreover, in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at the regional or province level,  and as far as the authors know, such data are not available.  For these reasons here the modeling will always be referred to as national data and national lock-downs. Therefore the lock-down that were applied starting October were not considered, while in the months of July, August and September the restriction were so weak that can be considered negligible for our model. Another point that is worth mentioning is that in this model, on the contrary concerning the SIS one, individuals can not re-infect: concerning the COVID-19, the possibility of reinfection is still discussed among scholars \cite{ledford2020covid,iwasaki2021reinfections}. As far as the authors know, it is established that the immunity lasts at least eight months  \cite{dan2021immunological} and very few reinfection cases are reported.  Thus the immunity considered in the model for one year seems almost a realistic approximation.

@@

The economic impact for each epidemiological scenario is shaped as follow: the first set of parameters,  as the economic value of death and of being infected by COVID-19, was taken directly from the reports/documentation of official sources; other parameters as the effect on sales for different areas and on unemployment benefit (Cassa Integrazione Guadagni) were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (300k EUR). This choice is based on the idea that,  if the PM misbehaves,  can be incriminated for manslaughter (with the consent of parliament that has to validate the incrimination) and then,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keep into account the age of the deaths: this in principle requires an epidemiological model in which also the age of people is taken into account.  However, as said before, a partial differential equations system should be solved, making the calculation and the computational cost incredibly high.}.  Besides this impact,  there is also the cost associated with the medical care of each ill people. For this, we considered the national average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another import health-care impact because the ill people for COVID-19 saturate the health system thus making it unreachable for other diseases.  This spillover translates into more death and more ill people for the baseline situation where there is not a pandemic: however,  by now,  this effect is difficult to quantify, and so we did not include it in the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the sales value for the following ATECO-2007 \cite{Ateco2007} categories \footnote{In the rest part of the paper, these categories will be referred to as the part of the name labeled in blue. 

@@

\textcolor{blue}{Photo-optics}/films/compact discs/audio-video cassettes and musical instruments, Durable and non-durable \textcolor{blue}{Homeware}, \textcolor{blue}{Household tools} and hardware tools, \textcolor{blue}{Games}/toys/sports and camping articles. The choice to use sales value as a parameter for the evaluation of the lock-down lies in the fact that with them is possible to capture not only the contraction for each sector but also the loss for the public treasury due to to the reduced incomes from the VAT\footnote{For this purpose another sector that in principle can also be considered is the contraction of fuel sales value, due to the reduced mobility, where in addition to the VAT there is also fixed taxation (accisa).  Such calculation may be considered as a future outlook of this work}.  Beside the sales value we also considered the unemployment benefit for the following ATECO sectors: \textcolor{blue}{Manufacturing} activities, \textcolor{blue}{Construction} , \textcolor{blue}{Wholesale} and retail trade/ repair of motor vehicles-motorcycles and personal and household goods. Furthermore also the total value (considering also other sectors that were not analysed here). In this case,  the choice to use also this parameter is based on the fact that this is the first aid provided by the Government for the firms that were damaged by the lock-down restrictions.  For the empirical evaluation of the impact on sales value and redundancy funds, we performed a multiple time Difference in Difference as presented in Refs.  \cite{draca2011panic},  \cite{imbens2009recent} and \cite{wooldridge2012introductory}. The following regression was performed: 

@@

where $Y$ is the selected outcome (sales value or redundancy funds), $\alpha$ the intercept, $T_{i}$ a dummy variable for the lock-down timing i,  $C$ a dummy for the treated group, and $\epsilon$ an error term.  As a control group for the sales value,  we considered the food ones since,  in principle,  people can be considered to use almost the same amount of food regardless of the lock-down\footnote{Although it is true that a slight increase of food sales value during the lock-down is present in the plots in Fig.  \ref{Selling_plots} \ref{Selling_plots_II},  it must be stressed the fact that, as proven numerically in the Supporting Info A, this change does not significantly affect the DID estimations }.  On the other side,  for redundancy funds \textit{Cassa integrazione},  we considered the Cassa Integrazione Straordinaria - Solidarietà as the control group (note that this subsidy is different respect respect Cassa Integrazione Solidarietà that was dedicated to sectors not covered by Cassa Integrazione Ordinaria).  This subsidy can be used by firms,  damaged by the pandemics and by the lock-downs,  in order to reduce their labor cost but at the same time guaranteeing to the workers part of their original salary \footnote{It is worth noting that the \textit{Cassa Integrazione Ordinaria} considered here is not the only one contribution that was provided by Italian Government: in fact, there was also, for instance,  the \textit{Cassa Integrazione in Deroga} that was dedicated to the firms not covered by the \textit{Ordinaria}.  However, in the present study, we consider, for simplicity,  only the sectors covered by Cassa Integrazione Ordinaria. }.  During the first months of the pandemic,  following the rules stated by the Italian Government (Decreto Cura Italia \cite{curaitalia}),  firms that would reduce their labor cost first forced the employees to use their holiday budget and then,  after it was run out,  they put the employees into the Cassa Integrazione Ordinaria.  Thus the Cassa Integrazione Straordinaria-Solidarietà can be considered as not treated by the first wave of the lock-down, while the ordinary one treated.  The DID consider here has three different times for the national lock-down:  for the sales value, we considered  the months of March 2020 and May 2020 as medium lock-down, the month of April 2020 as high lock-down, and June 2020 as low lock-down.  On the other side concerning the redundancy funds,  we considered as medium lock-down only May 2020 while for the high and low lock-down the timing were identical to the sales value. Such choice was motivated by the fact that the effect of firms to use the redundancy funds for workers was slightly delayed because in March 2020 firms forced their employees first to use their holiday and then the \textit{Cassa Integrazione Ordinaria}: as a consequence,  because of this buffer effect,  in March there is no significant effect of this redundancy funds although there is a significant reduction of hours worked (see, e.g., \cite{istat_lavoro}).  Performing the DID, we estimated the coefficients for each intensity of lock-down on sales value and redundancy funds. Then these were rescaled in order to obtain a weekly value.  Finally, we multiplied the coefficients obtained from the DID (scaled from months to weeks) by the number of lock-down weeks,  with the respective intensity,  in each scenario simulated via the epidemiological model described before.  As a consequence, we obtained,  for each scenario,  an economics simulation based on parameters obtained from an empirical evaluation.

@@

The monthly sales value data (with the baseline $100=2015$) were retrieved from the National Institute of Statistics website \cite{istat}.  In particular, we considered the period starting from June 2018 up to June 2020 for the eight sales value categories described in the previous section . The choice to not consider the months after June and, in particular, the last part of the year lies in the fact that in the latter lock-downs were imposed at the regional level and not at national level \footnote{With the only exception of the Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed on regional cluster making it more flexible: unfortunately the economic data used here, as far as the author knows,  were not available,  at all for a regional cluster.  Moreover, if regional clusters were considered, it was necessary to model an ensemble of SIRD models that communicate with a defined rate (that change also with respect to the lock-down restriction).  This makes the model much more complicated}.  As discussed in Supporting Information,  these data were not de-seasoned. Thus, we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transformation \cite{huang1998empirical} for the decomposition of a time series data.  The decomposition results are reported in the Supporting Info E.  Concerning monthly redundancy funds paid we retrieved the data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}. Here the period considered also starts from June 2018 up to June 2020 for the four categories described in the previous section.  It is worth noting that we considered only the authorized (paid) redundancy funds and not the asked ones.   Furthermore,  differently for the sales value, these data were not affected by seasonality noise,  and thus no de-seasoning was necessary.

@@

As done for the section Model and methods, we will divide the discussion of the results in the following way: first, the outcomes of the economic model will be presented,  then basing on this result,  we will discuss the scenarios obtained with them via the epidemiological model.  Finally, we will discuss the overall results.

@@

For each sales value category we run the regression reported in the Eq.  \ref{DID-eq} on de-seasoned data reported in Fig \ref{Selling_plots} and \ref{Selling_plots_II}.  The coefficients obtained are reported in Tab.  \ref{selling_DID_table_delta} (the estimation of the other parameters is given in Supporting Information D).  As one can point out from the plots, there is indeed a pre-trend in the data before the event,  however as proved numerically in the Supporting Info B,  the slope of this pre-trend is up to two orders of magnitude lesser for the slope in the lock-down T$_{1}$ and $T_{2}$.  Consequently, this pre-trend,  compared to the lock-down effect,  can be considered negligible.   As a further check,  a placebo test was performed by choosing timing before the pandemic of COVID19.  As illustrated in Supporting Info C,  this test was successful.  On the other side,  concerning $T_{3}$  (low lock-down),  only the effect for clothing and footwear can be considered significant.  Thus,  concerning $T_{3}$,  we considered not null,  in the scenario simulations,  only the clothing, and footwear sectors. The same DID regression was used for the redundancy funds,  using as the control group the extraordinary solidarity redundancy funds (cassa integrazione straordinaria solidarietà): the plot of the data and the DID coefficients are reported in Fig. \ref{CIG_sector_plots} and Tab \ref{CIG_DID}.  Also in this case the placebo tests were successfully performed.  An interesting insight is also provided by the inspection of the extraordinary redundancy funds for renovation ( \textit{cassa integrazione ristrutturazione}): these redundancy funds, compared to the solidarity ones, seem much more sensitive for an external shock in particular for the trade sector.

>>>>

Within a standard compartmental model for describing the dynamics of epidemic (Susceptible-Infectious-Recovered-Dead),  we considered a policy-maker (PM) that imposes stochastically different types of lock-downs. The probability that tunes this stochastic process reflects his/her different attitude to face an epidemics (e.g., \textit{laissez-faire} vs. very strict).  In order to simulate not only an epidemiological scenario but also an economic one, we estimated, via a Difference-in-Difference regression,  the impact of national lock-downs applied during the COVID19 first wave in Italy from March 2020 to June 2020,  on two microeconomic sectors: sales values and redundancy funds (\textit{Cassa Integrazione}).  We found that by modifying with continuity the PM attitude to impose the lock-down, a phase transition (as defined for a physical system) is obtained.  The comparison of these two scenarios and their impact provides a bird's-eye view of the socio-economic consequences of the PM attitude

@@

The recent pandemic due to the spread of the SARS-CoV-2 virus opened a highly debated issue about the best approach for the policy-maker to face the epidemic.  Unlike past pandemics of XX century (e.g., Spanish Flu,  Asiatic Flu, and Hong-Kong Flu), a massive amount of data are easily accessible for this pandemic.  Consequently, the modeling of the virus diffusion and the effect on socio-economic texture for different countries can be investigated with a more satisfactory resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally, the reproduction number, as described in Supporting Information).  Indeed the policy-maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme there is a slowdown in the spread of the virus but with a consequence not only on economic activity (and so income of people/firms) but also that the furniture of many primary goods is stopped.  On the other side,  if the PM takes no lock-down measures, the toll to be paid will be not only the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice, the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion than others (for this purpose, an excellent analysis was provided by Li et al. in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).
As a consequence, the lock-down efficacy,  within certain limits,  can be regulated.  In the literature, different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular, for the model in Ref. \cite{kabir2020evolutionary}, it is assumed that the PM can impose a continuous factor for the lock-down (e.g., he/she can reduce the daily number of contacts between people by choosing each value between a definite range). In contrast, for different countries, such a factor seems to be much more discrete (e.g., the PM can reduce the daily contacts only by choosing definite values). Moreover it is assumed that the policy-maker perfectly knows the consequences of his/her choices and that he/she can act without delay to impose the optimal choice: this feature is entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy-maker.  Here we would propose an alternative way to model the decision of the policy-maker that is based on a stochastic model instead of on a deterministic one. Moreover we will consider a discrete intensity for the lock-down. Finally,  differently to previous researches that focused basically on a macroeconomic impact,  here we estimated and put in the model the impact of the different lock-downs at the microeconomic level. In particular, through difference in difference (DID) ,  we evaluated the effect of the different levels of lock-down on different sales sectors as well as on redundancy funds (\textit{Cassa Integrazione Ordinaria}) in Italy.  Thus, the final output of the model will be the cumulative deaths, the economic damage for each sale sector and the increase of redundancy funds paid.
Here we will also consider that there is not only an economic cost for each death as done by \cite{kabir2020evolutionary}, but there is also an average cost for each infected person (referring to Italian data) because a consistent part of them should be assisted by the national health system.  As we will show by varying with continuity the probabilistic parameter by which the PM imposes the lock-down,  a discontinuity in the epidemiological model was obtained.  Such behaviour belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase, we will first discuss the result of the simulation, then we will compare them to get a general insight.

\section{Models and methods}

The model of this study is composed of an epidemiological part that shapes the diffusion of the virus. Its output is then used by the economic model to quantify the damage. Thus we will discuss the epidemiological part and then the economic one

@@

where S is the number of people that are still susceptible, I the number of people that are infected, R people that are recovered, and D people that are dead.  N denotes the total population that for the timing of this paper will be considered fixed\footnote{Otherwise if longer horizontal timing is considered, it is necessary to consider a source term for the births and a well term for the natural deaths. For further details see \cite{vynnycky2010introduction}}. On the other side, $\beta$,$\gamma$, and $\mu$ are the parameters that shape the probability by which one individual in the model moves from a compartment to another: in particular, $\beta$ is the probability to be infected, $\gamma $ the probability to recover and $\mu$ the probability to die (basically the lethality that is defined as the probability to die given to be ill).  Usually, epidemiologists are interested in the ratio:  

@@

known as the basic reproduction factor. This number is the average number of people that a single individual infects and describes if the epidemic is in negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in positive feedback (R$_{0}$>1).  As a consequence, if the epidemic is within a negative feedback will be dissipated, while if it is in positive feedback will grow.  Note that in this simple model, since the parameters are not time-dependent, this factor is constant.  As performed by Ferrari et al. \cite{ferrari2020modelling},  when time-dependent parameters $\beta$ and $\gamma$ are taken in to account the reproduction factor R$_{0}$ becomes time dependent: thus scholar rename it as $R_{t}$.  For the present work,  we limited to constant parameters. In particular, we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$. $\mu$ was set in order to keep into account the calculated lethality for Italy: 1 \% \cite{fernandez2020estimating} (see also the excellent analysis made by the Institute for International Political Studies (ISPI) \cite{ISPI})\footnote{The authors are aware that this ratio is far from being homogeneous for the different ages of the population: however if this factor is taken into account,  it requires the solution of a system of partial derivative equations. In this case, the numerical calculations become much more complicated}.  The overall population $N$ was set to 60M to simulate the Italian population.  Within the daily temporal evolution of this model, which was obtained by numerically solving the differential equation above via the \textit{DeSolve} package, we considered a trigger activated by the PM every seven days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM imposes laws that reduces $\beta$ factor by a multiplicative factor equal to $0.7$ (and thus the reproduction factor $R_{0}$). If the normalized infected people are more than ten times the previous threshold he/she will impose with certain probability restrictions that reduce $\beta$ factor by a multiplicative factor of $0.25$; finally if the threshold is exceeded more than 50 times, the PM will impose with a certain probability restrictive measures that reduce $\beta$ to a multiplicative factor equal to $0.025$. These attenuation parameters were adjusted, considering results of Marziano et al. in Ref. \cite{marziano2021retrospective}.  Therefore such a trigger makes the  R$_{0}$ parameter time-dependent,  although in a discrete way.  As we said, the PM acts with a certain probability,  more formally stochastically; each week, a random number (from zero to one) is extracted: if this is higher than a certain threshold,  the relative restrictive decision is taken, otherwise not. The threshold value captures the PM attitude to impose the lock-down: lower values model a careful PM, high values a lazy one. In this way, the model can simulate different scenarios for the different PM attitudes: as we will see, this can produce two very different results.  At the end of the simulation, besides the values given by the standard SIRD model (recovered and deaths),  the algorithm also provides the number of weeks in which each restriction was active. These values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy. It is worth noting that here as lock-down, we considered only the national one applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modeling a unique system is easier with respect to modeling an ensemble of communicating clusters that represent regions or provinces: therefore, if one is interested in modeling the second wave of pandemics, such an approach should be undertaken. Moreover, in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at the regional or province level and as far as the authors know, such data are not available.  For these reasons here the modeling will always be referred to as national data and national lock-downs. Therefore the lock-down that were applied starting October were not considered, while in the months of July, August and September the restrictions were so weak that can be considered negligible for our model. Contrary to the SIS (susceptible-infected-susceptible) one, individuals can not re-infect: concerning the COVID-19, the possibility of reinfection is still discussed among scholars \cite{ledford2020covid,iwasaki2021reinfections}. As far as the authors know, it is established that immunity lasts at least eight months  \cite{dan2021immunological} and very few reinfection cases are reported.  Thus the year immunity considered in the model seems almost a realistic approximation.

@@

The economic impact for each epidemiological scenario is shaped as follow: the first set of parameters,  as the economic value of death and of being infected by COVID-19, was taken directly from the reports/documentation of official sources; other parameters as the effect on sales for different areas and on redundancy funds (Cassa Integrazione Guadagni), were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (332k EUR). This choice is based on the idea that,  if the PM misbehaves,  can be incriminated for manslaughter (with the consent of the parliament that has to validate the incrimination) and then,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keeps into account the age of the deaths: this in principle requires an epidemiological model in which also the age of people is taken into account.  However, as said before, a partial differential equations system should be solved, making the calculation and the computational cost incredibly high.}.  Besides this impact,  there is also the cost associated with the medical care of each ill people. For this we considered the national average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another important health-care impact because COVID-19 ill people saturate the health system making it unreachable for other diseases.  This spillover translates into more deaths and more ill people for the baseline situation where there is not a pandemic: however,  by now,  this effect is difficult to quantify and so we did not include it in the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the sales values for the following ATECO-2007 \cite{Ateco2007} categories \footnote{In the remaining part of the paper, these categories will be referred to as part of the name labelled in blue. 

@@

\textcolor{blue}{Photo-optics}/films/compact discs/audio-video cassettes and musical instruments, Durable and non-durable \textcolor{blue}{Homeware}, \textcolor{blue}{Household tools} and hardware tools, \textcolor{blue}{Games}/toys/sports and camping articles. The choice to use sales values as a parameter for the evaluation of the lock-down lies in the fact that with them is possible to capture not only the contraction for each sector but also the loss for the public treasury due to the reduced incomes from the VAT\footnote{For this purpose another sector that can also be considered is the contraction of fuel sales values, due to the reduced mobility, where in addition to the VAT there is also fixed taxation (accisa).  Such calculation may be considered as a future outlook of this work}.  Beside the sales values we also considered the redundancy funds for the following ATECO sectors: \textcolor{blue}{Manufacturing} activities, \textcolor{blue}{Construction} , \textcolor{blue}{Wholesale} and retail trade/ repair of motor vehicles-motorcycles and personal and household goods. Furthermore, also the total value was included (its value is given also by the sum of other sectors that were not analysed here). In this case, the choice to consider the redundancy funds is based on the fact that this is the first aid provided by the Government for firms that were damaged by lock-down restrictions.  For the empirical evaluation of the impact on sales values and redundancy funds, we performed a multiple time Difference in Difference as presented in Refs.  \cite{draca2011panic},  \cite{imbens2009recent} and \cite{wooldridge2012introductory}. The following regression was performed: 

@@

where $Y$ is the selected outcome (sales values or redundancy funds), $\alpha$ the intercept, $T_{i}$ a dummy variable for the lock-down timing i,  $C$ a dummy for the treated group, and $\epsilon$ an error term. As a control group for sales values we considered food since people almost consumed the same amount despite the lock-down\footnote{Although it is true that a slight increase of food sales values during the lock-down is present in the plots in Fig.  \ref{Selling_plots} \ref{Selling_plots_II},  it must be stressed the fact that, as proven numerically in the Supporting Info A, this change does not significantly affect the DID estimations }.  On the other side,  for redundancy funds \textit{Cassa integrazione},  we considered the Cassa Integrazione Straordinaria - Solidarietà as control group (note that this subsidy is different respect respect Cassa Integrazione Solidarietà that was dedicated to sectors not covered by Cassa Integrazione Ordinaria).  This subsidy can be used by firms damaged by the pandemics and by the lock-downs  in order to reduce their labor cost but at the same time guaranteeing to the workers part of their original salary \footnote{It is worth noting that the \textit{Cassa Integrazione Ordinaria} considered here is not the only one contribution that was provided by Italian Government: in fact, there was also, for instance,  the \textit{Cassa Integrazione in Deroga} that was dedicated to the firms not covered by the \textit{Ordinaria}.  However, in the present study, we consider, for simplicity,  only the sectors covered by Cassa Integrazione Ordinaria. }.  During the first few months of the pandemic,  following the rules stated by the Italian Government (Decreto Cura Italia \cite{curaitalia}),  firms that reduces their labor cost first forcing the employees to use their holiday budget and then,  after consuming it all,  they put the employees into Cassa Integrazione Ordinaria.  Thus the Cassa Integrazione Straordinaria-Solidarietà can be considered as not treated by the first wave of the lock-down, while the ordinary one treated. DID considered here has three different time period for the national lock-down:  for the sales values, we considered  the months of March 2020 and May 2020 as medium lock-down, the month of April 2020 as high lock-down, and June 2020 as low lock-down.  On the other side concerning redundancy funds we considered as medium lock-down only May 2020 while for the high and low lock-down the timing were identical to the sales values. Such choice was motivated by the fact that the effect of firms to use the redundancy funds for workers was slightly delayed because in March 2020 firms forced their employees first to use their holiday and then the \textit{Cassa Integrazione Ordinaria}. As a consequence,  because of this buffer effect,  in March there is no significant effect of this redundancy funds although there is a significant reduction of hours worked (see, e.g., \cite{istat_lavoro}).  Performing the DID we estimated the coefficients for each intensity of lock-down on sales values and redundancy funds. These values were then rescaled in order to obtain weekly value.  Finally, we multiplied the coefficients obtained from the DID (scaled from months to weeks) by the number of lock-down weeks,  with the respective intensity, for each scenario simulated via the epidemiological model described before.  As a consequence we obtained,  for each scenario,  an economics simulation based on parameters obtained from an empirical evaluation.

@@

The monthly sales values data (with the baseline $100=2015$) were retrieved from the National Institute of Statistics website \cite{istat}.  In particular, we considered the period starting from June 2018 up to June 2020 for the eight sales values categories described in the previous section . The choice to not consider the months after June and, in particular, the last part of the year lies in the fact that in the latter lock-downs were imposed at a regional level and not at a national level \footnote{With the only exception of the Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed at a regional cluster making it more flexible: unfortunately the economic data used here, as far as the author knows,  were not available,  at all for a regional cluster.  Moreover, if regional clusters were considered, it was necessary to model an ensemble of SIRD models that communicate with a defined rate (that change also with respect to the lock-down restriction).  This makes the model much more complicated}.  As discussed in Supporting Information,  these data were not de-seasoned. Thus, we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transformation \cite{huang1998empirical} for the decomposition of a time series data.  Decomposition results are reported in the Supporting Info E.  Concerning monthly redundancy funds paid we retrieved data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}. Here the period considered also starts from June 2018 up to June 2020 for the four categories described in the previous section.  It is worth noting that we considered only the authorized (paid) redundancy funds and not the asked ones.   Furthermore,  differently from the sales values, these data were not affected by seasonality noise,  and thus no de-seasoning was necessary.

@@

As done for the section \textit{Model and methods}, we will divide  discussion of the results in the following way: first, the outcomes of the economic model will be presented then, based on this result,  we will discuss the scenarios obtained with them via the epidemiological model.  Finally, we will discuss the overall results.

@@

For each sales values category we run the regression reported in Eq.  \ref{DID-eq} on de-seasoned data reported in Fig \ref{Selling_plots} and \ref{Selling_plots_II}.  The coefficients obtained are reported in Tab.  \ref{selling_DID_table_delta} (estimation of the other parameters is given in Supporting Information D).  As one can point out from the plots, there is indeed a pre-trend in the data before the event,  however as proved numerically in the Supporting Info B,  the slope of this pre-trend is up to two orders of magnitude lesser for the slope in the lock-down T$_{1}$ and $T_{2}$.  Consequently, this pre-trend,  compared to the lock-down effect,  can be considered negligible.   As further check, placebo test was performed by choosing timing before the COVID19 pandemic.  As illustrated in Supporting Info C,  this test was successful.  On the other side,  concerning $T_{3}$  (low lock-down),  only the effect for clothing and footwear can be considered significant.  Thus,  concerning $T_{3}$,  we considered not null,  in the scenario simulations,  only clothing and footwear sectors. The same DID regression was used for the redundancy funds,  using as control group the extraordinary solidarity redundancy funds (cassa integrazione straordinaria solidarietà): the plot of the data and the DID coefficients are reported in Fig. \ref{CIG_sector_plots} and Tab \ref{CIG_DID}.  Also in this case the placebo tests were successfully performed.  An interesting insight is also provided by the inspection of the extraordinary redundancy funds for renovation ( \textit{cassa integrazione ristrutturazione}): these redundancy funds, compared to the solidarity ones, seem much more sensitive for an external shock and in particular for the trade sector.
